Assessing landscape N removal in coastal New England catchments using the

### 2023-10-25 manuscript/manuscript.Rmd
Small changes in preparation for submission to f1000Research

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.^*3*^, Lei-Parent,
  Qian^*3*^, Forshay, K. J.^*4*^"
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^corresponding author: <qkellogg@uri.edu>'
bibliography: manuscript.bib
output:
  word_document:
    keep_md: true
    reference_docx: manuscript_template.docx
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features frequently found within the headwaters
  of watersheds can act as landscape sinks for N, such as wetlands, streams and ponds
  that transform dissolved N into gaseous N, effectively removing it from the aquatic
  system. Identifying and evaluating these landscape sinks and their spatial relationship
  to N sources can help managers understand the effects of alternative decisions on
  downstream resources. N-Sink is an approach that uses widely available GIS data
  to identify landscape sinks within HUC-12 (or larger) catchments, estimate their
  N removal potential and map the effect of those sinks on N movement through the
  catchment. A series of static maps are produced to visualize N removal efficiency,
  transport and delivery, the latter in the form of an index. To facilitate data acquisition,
  processing and visualization, an R package `nsink` was developed and applied to
  produce static maps for all HUC-12 catchments in southern New England. The R package
  can be used to develop N delivery maps for a specific HUC-12 of interest, or users
  can visit the University of Connecticut website to explore previously mapped areas.
  Users can also investigate specific flowpaths interactively by clicking on any location
  within the catchment. A flowpath is generated, along with a table describing N removal
  along each segment."
---

@@

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]), and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. The approach relies on residence time within landscape features that support N removal as the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

## Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD data (e.g. time of travel) associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system. We examined seven HUC-12s within southern New England, with a range of land cover and soils, for the extent of off-network hydrology. None had off network stream or canal segments, while six of the seven had off-network waterbodies. These off-network waterbodies represented from 2 to 14% (mean of 7%) of total lake area. We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. A subset of these off-network stream segments are identified by the NHD dataset as ditches and canals. These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment due to the level of uncertainty regarding residence time in these groundwater-fed lakes and ponds [@hampton2019residence].

## Caveats
N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated. In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage are present, flow can bypass the removal processes within wetland ecosystems [@gold2001landscape]. In these situations first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated. N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin. Here, nitrogen-enriched groundwater may enter estuaries directly as submarine discharge rather than via baseflow to stream networks [@nowicki2008groundwater], requiring site specific assessments to evaluate N removal in salt marshes or as upwelling into the seabed [@addy2005denitrification;@robinson2018groundwater]. These situations have been estimated to account for approximately 6% of freshwater inputs to estuaries globally and in areas of New England [@burnett2003groundwater;@nowicki2008groundwater].

@@

Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters. This example illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Figure 4). The difference is due to the different N sinks encountered along the way to the estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3). These two currently undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet. If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky "eastern region (flowpath B) underlines the importance of wetlands and slow moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.

@@

Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps. The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment. Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.

>>>>

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with the R package, nsink"
author:
- "Kellogg, D. Q.^\\*^^*1*^, Hollister. J. W. ^*2*^, Arnold, C. L.^*3*^, Gold, A.
  J. ^*1*^, Wilson, E. H.^*3*^, Chadwick, C. B.^*3*^, Dickson, D. W.^*3*^, Lei-Parent,
  Qian^*3*^, Forshay, K. J.^*4*^"
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^corresponding author: <qkellogg@uri.edu>'
bibliography: manuscript.bib
output:
  word_document:
    keep_md: true
    reference_docx: manuscript_template.docx
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features frequently found within the headwaters
  of watersheds can act as landscape sinks for N, such as wetlands, streams and ponds
  that transform dissolved N into gaseous N, effectively removing it from the aquatic
  system. Identifying and evaluating these landscape sinks and their spatial relationship
  to N sources can help managers understand the effects of alternative decisions on
  downstream resources. N-Sink is an approach that uses widely available GIS data
  to identify landscape sinks within HUC-12 (or larger) catchments, estimate their
  N removal potential and map the effect of those sinks on N movement through the
  catchment. A series of static maps are produced to visualize N removal efficiency,
  transport and delivery, the latter in the form of an index. To facilitate data acquisition,
  processing and visualization, an R package `nsink` was developed and applied to
  produce static maps for all HUC-12 catchments in southern New England. The R package
  can be used to develop N delivery maps for a specific HUC-12 of interest, or users
  can visit the University of Connecticut website to explore previously mapped areas.
  Users can also investigate specific flowpaths interactively by clicking on any location
  within the catchment. A flowpath is generated, along with a table describing N removal
  along each segment.In this software tool article we describe the motivation behind developing 
  nsink, discuss the implementation in R, and describe the use of nsink with two use case examples. 
  nsink is available from <https://github.com/USEPA/nsink>."
---

@@

N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale (HUC = Hydrologic Unit Code; [@berelson2004mapping]) and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in Kellogg et al. [-@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [@mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. The approach relies on residence time within landscape features that support N removal as the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

## Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD data (e.g. time of travel) associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. For the vast majority of HUC-12 catchments off-network features are rare and spatially small compared to the overall system. We examined seven HUC-12s within southern New England, with a range of land cover and soils, for the extent of off-network hydrology. None had off network stream or canal segments, while six of the seven had off-network waterbodies. These off-network waterbodies represented from 2 to 14% (mean of 7%) of total lake area. We estimated removal from off-network stream segments to be equal to the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. A subset of these off-network stream segments is identified by the NHD dataset as ditches and canals. These are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore N-Sink assigns removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment due to the level of uncertainty regarding residence time in these groundwater-fed lakes and ponds [@hampton2019residence].

## Caveats
N-Sink was designed for catchments where surface and groundwater flow paths are not highly manipulated. In settings where substantial storm water conveyance networks or extensive subsurface agricultural drainage are present, flow can bypass the removal processes within wetland ecosystems [@gold2001landscape]. In these situations, first-order streams may be channelized, leading to higher velocities than in undisturbed channels and the N-Sink estimates of removal may be overstated. N-Sink does not account for nitrogen removal from high permeability (e.g., sand and gravel) unconfined aquifers along the coastal margin. Here, nitrogen-enriched groundwater may enter estuaries directly as submarine discharge rather than via baseflow to stream networks [@nowicki2008groundwater], requiring site specific assessments to evaluate N removal in salt marshes or as upwelling into the seabed [@addy2005denitrification;@robinson2018groundwater]. These situations have been estimated to account for approximately 6% of freshwater inputs to estuaries globally and in areas of New England [@burnett2003groundwater;@nowicki2008groundwater].

@@

Using this approach, we can compare the N delivery from two locations that are separated by approximately 900 meters. This example illustrates the importance of location with respect to landscape N sinks. Approximately 56% of the N that originates at point A is transported to the estuary vs. approximately 25% of the N that originates at point B (Figure 4). The difference is due to the different N sinks encountered along the way to the estuary. Most of the removal that occurs along flowpath A is due to flow through hydric (wetland) soils, with little removal occurring during flow along the downgradient stream network (Table 2). Flowpath B encounters a large pond along the way, significantly increasing residence time and N transformation (Table 3). These two currently undeveloped locations demonstrate the importance of understanding the landscape position and its role in the movement of N from source to outlet. If development were proposed in the western portion of this area (flowpath A), source controls could be advised, while the less "leaky" eastern region (flowpath B) underlines the importance of wetlands and slow-moving water ways, such as ponds, that protect downstream waters by slowing flow and increasing opportunities for N transformation.

@@

Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps. The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier" parts of a catchment. Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.
